---
#title: "Milestone 2, Group 11"
format: pdf
editor: visual
---

```{r include=FALSE, message=FALSE, warning=FALSE}

#install.packages("readxl")
library(readxl)
library(janitor)
library(dplyr)
library(knitr)
library(readr)
library(tidyverse)
#install.packages("tinytex")
#tinytex::install_tinytex()
library(gt)
```

```{r, include=FALSE}

ces_datadictionary <- read_excel("~/PHW251_2025/251-Group11/calenviroscreen_datadictionary.xlsx")
  
head(ces_datadictionary)
```

```{r, include=FALSE}
ces_measures <- read.csv("~/PHW251_2025/251-Group11/calenviroscreen_measures_2021.csv")

ces_measures <- clean_names(ces_measures) 

ces_measures <- ces_measures %>% mutate(census_tract = as.character(census_tract))

#this is to show the type of each element
str(ces_measures%>% select(california_county, asthma, pollution_burden_score, pm2_5, diesel_pm, poverty))

head(ces_measures)
```

```{r, include=FALSE}
#showing the summary statistics
summary(ces_measures$asthma)

```

```{r, include=FALSE}

unique(ces_measures$california_county)
```

```{r, include=FALSE}
summary(ces_measures$poverty)
```

```{r, include=FALSE}
summary(ces_measures$diesel_pm)
```

```{r, include=FALSE}
summary(ces_measures$pm2_5)
```

```{r, include=FALSE}
summary(ces_measures$pollution_burden_score)
```

```{r, include=FALSE}
ces_scores_demog <- read.csv("~/PHW251_2025/251-Group11/calenviroscreen_scores_demog_2021.csv")

ces_scores_demog <- clean_names(ces_scores_demog)

head(ces_scores_demog)
```

```{r, include=FALSE}
chhs_asthma_ed <- read.csv("~/PHW251_2025/251-Group11/chhs_asthma_ed.csv")

chhs_asthma_ed <- clean_names(chhs_asthma_ed)
str(chhs_asthma_ed%>% select(county, number_of_ed_visits, age_adjusted_ed_visit_rate))


head(chhs_asthma_ed)
```

```{r, include=FALSE}
summary(chhs_asthma_ed$number_of_ed_visits)
```

|     |
|-----|
|     |

|     |     |
|-----|-----|
|     |     |

```{r, include=FALSE}
#angela asthma dataframe 
df_asthma <- chhs_asthma_ed %>% 
  clean_names() %>%
  filter(county != "CALIFORNIA") %>% 
  filter(strata == "Total population") %>% 
  filter(year == 2020) %>% 
  select(county, year, number_of_ed_visits, age_adjusted_ed_visit_rate) %>%
  mutate(county = str_to_title(county)) %>%     
  mutate(county = paste0(county, " County")) %>%     
  mutate(across(where(is.numeric), ~if_else(is.na(.), 0, .)))
df_asthma
```

```{r, include=FALSE}
#dineli environmental measures dataset clean up
df_env_measures <- read_csv("calenviroscreen_measures_2021.csv", show_col_types = FALSE) %>%
  clean_names() %>%
  
  # recode common variable(s) for combining with other datasets
  mutate(county = str_to_title(california_county)) %>%
  mutate(county = paste0(county, " County")) %>%
  
  # recode values that are reading incorrectly
  mutate(across(where(is.numeric), ~na_if(., -999))) %>%
  mutate(across(where(is.numeric), ~na_if(., -1))) %>%
  
  # identify and select environmental measures of interest
  select(county, pm2_5, diesel_pm, poverty, traffic) %>%
  
  # group by county and summarize to one row per county (median)
  group_by(county) %>%
  summarise(
    median_pm2_5 = median(pm2_5, na.rm = TRUE),
    median_diesel_pm = median(diesel_pm, na.rm = TRUE),
    median_poverty = median(poverty, na.rm = TRUE),
    median_traffic = median(traffic, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # replace missing numeric values with 0 (to match your asthma dataset syntax)
  mutate(across(where(is.numeric), ~if_else(is.na(.), 0, .)))

df_env_measures 

```

```         
```

```{r, include=FALSE}

#angela CES scores data frame

df_scores <- ces_scores_demog %>%
  group_by(county) %>%
  summarise(
    mean_ces_4_0_score = mean(ces_4_0_score, na.rm = TRUE),
    total_population = sum(total_population, na.rm = TRUE),
    .groups = "drop"
  )
df_scores

```

## <!-- ## Combine Dataframes and Re-Structured -->

```{r, include=FALSE}

df_combined <- df_scores %>%
  left_join(df_asthma, by = "county") %>%
  left_join(df_env_measures, by = "county") %>%
  mutate(ed_visits_per_100k = (number_of_ed_visits /total_population) * 100000)

df_combined


```

## <!-- ## Data re-structured and summarized -->

```{r, include=FALSE}
df_combined %>%
  dplyr::summarise(
    n_counties = dplyr::n(),
    pm25_med   = median(median_pm2_5, na.rm = TRUE),
    diesel_med = median(median_diesel_pm, na.rm = TRUE),
    poverty_med= median(median_poverty, na.rm = TRUE),
    ces_mean   = mean(mean_ces_4_0_score, na.rm = TRUE),
    asthma_rate_med = median(age_adjusted_ed_visit_rate, na.rm = TRUE)
  ) %>%
  gt() %>%
  tab_header(title = "df_combined — Summary Statistics") %>%
  fmt_number(columns = where(is.numeric), decimals = 2)

```

## <!-- ## Create a Table for HTML Output -->

```{r, include=FALSE}
library(gt)
#Create a clean table for html output
df_combined %>%
  
  gt() %>%
  tab_header(
    title = "Merged County-Level Dataset",
    subtitle = "CalEnviroScreen Scores, Environmental Indicators, and Asthma ED Visits (2020)"
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 2
  ) %>%
  tab_options(
    table.width = pct(100),
    table.font.size = px(13)
  )

```

```         
```

## <!-- ## Data Dictionary -->

```{r, include=FALSE}
#Data Dictionary

library(gt)

data_dictionary <- tibble::tibble(
  variable = c(
    "county",
    "mean_ces_4_0_score",
    "median_pm2_5",
    "median_diesel_pm",
    "median_poverty",
    "median_traffic",
    "number_of_ed_visits",
    "ed_visits_per_100k"
  ),
  type = c(
    "character",
    "numeric","numeric","numeric","numeric","numeric",
    "numeric","numeric"
  ),
  description = c(
    "County name.",
    "County mean of tract-level CalEnviroScreen 4.0 score (CES score = Pollution Burden × Population Characteristics).",
    "County median of tract-level annual average PM2.5 (µg/m³).",
    "County median of tract-level Diesel PM concentration surrogate/index (from on- and off-road sources).",
    "County median of tract-level percent of residents below 200% of the federal poverty level.",
    "County median of tract-level traffic density (vehicle-km per hour per road length) within 150 m of tract boundary.",
    "Total asthma-related ED visits (count) for the selected year (e.g., 2020), Total population, All ages.",
    "Derived rate = (number_of_ed_visits / total_population) × 100,000; NA if population not available."
  )
)

data_dictionary %>%
  gt() %>%
  tab_header(title = "Data Dictionary — Clean County-Level Dataset")
```

# MILESTONE 4 STARTS HERE 

Please submit an Rmd or Qmd and publish an html file on RPubs with the following:

1.  Final datasets for creation of visualization

    -   Join all datasets together

    -   Calculate any remaining data elements needed for analysis

    -   Show code used to create joined dataset, but please do not print full data frame output (showing data structure with `str()` is okay)

2.  Visualizations (at least one per group member)

    -   One print quality table as requested in scenario

    -   One print quality plot or chart as requested in scenario

    -   For groups of 3, one additional print quality table or plot of your choice (can support the requested data in the scenario, or answer a different question using the same data sources)

3.  For each visual, include

    -   Code used to generate visual

    -   Legend (if necessary)

    -   1-2 sentence interpretation

    -   NOTE:

        -   Please make sure the visual can stand-alone, meaning it includes enough information in title, legend, and footnote so that a person who sees only the visualization could understand what is being presented.

        -   Please also make sure column names, axis labels, and any other labels are meaningful and not just the name of the variable (ex: "County" rather than "county_name")

4.  Html file that is professionally prepared for presentation and published to RPubs

    -   Only the necessary information is in the output (e.g., suppress entire data frame outputs but showing data structure with `str()` is okay.

    -   Code used to import and clean data (milestones 2 and 3) can be included in the rmd/qmd but please use the option include=FALSE in order to prevent the code and any output from. Code chunk would look like `{r chunk_name, include=FALSE}`{style="font-family: Monaco, Menlo, Consolas, \"Courier New\", monospace; font-size: 0.75rem; border-radius: 6px; background-color: rgb(242, 244, 244); border: 1px solid rgb(232, 234, 236); padding: 0.125rem 0.25rem; color: rgb(199, 31, 35);"}

    -   Utilize `warning=FALSE`{style="font-family: Monaco, Menlo, Consolas, \"Courier New\", monospace; font-size: 0.75rem; border-radius: 6px; background-color: rgb(242, 244, 244); border: 1px solid rgb(232, 234, 236); padding: 0.125rem 0.25rem; color: rgb(199, 31, 35);"} and `message=FALSE`{style="font-family: Monaco, Menlo, Consolas, \"Courier New\", monospace; font-size: 0.75rem; border-radius: 6px; background-color: rgb(242, 244, 244); border: 1px solid rgb(232, 234, 236); padding: 0.125rem 0.25rem; color: rgb(199, 31, 35);"} to suppress any warnings/messages that may be displaying from any chunks

    -   For this milestone please make sure code for visualizations are included in the final file. Show your work by "echoing" code used in Rmd/Qmd file to create your tables and graphs. \`\`\`{r, echo = TRUE} in code chunk preferences.

    -   Use headers and sub headers to create an organized document

    -   Make sure you commit and push your rsconnect folder! [Read more details here.Links to an external site.](https://docs.google.com/document/d/1JSg5JUEI3JDD_rPAlAaVPno0MlUGQ6W_cyBvn6Ud1xs/edit?usp=sharing)

# Final datasets for creation of visualization

```{r 1, echo = TRUE}
# read in and clean all 3 files (reuse from Milestone 3)

# chhs_asthma_ed, calenviroscreen_measures_2021, ces_scores_demog

# ... your existing df_asthma, df_env_measures, df_scores code ...

df_combined <- df_scores %>%
left_join(df_asthma,      by = "county") %>%
left_join(df_env_measures, by = "county")

# any extra derived variables (env_exposure_index, ces_category, etc.)

# e.g.:

df_combined <- df_combined %>%
mutate(
ces_category = case_when(
mean_ces_4_0_score >= quantile(mean_ces_4_0_score, 2/3, na.rm = TRUE) ~ "High",
mean_ces_4_0_score <= quantile(mean_ces_4_0_score, 1/3, na.rm = TRUE) ~ "Low",
TRUE ~ "Moderate"
)
)

str(df_combined)
```

```{r}
#
df_combined %>%
select(
county,
mean_ces_4_0_score,
median_pm2_5,
median_poverty,
age_adjusted_ed_visit_rate
) %>%
arrange(desc(age_adjusted_ed_visit_rate)) %>%
slice_head(n = 10) %>%
gt() %>%
tab_header(
title = "Counties with Highest Asthma ED Visit Rates",
subtitle = "With CalEnviroScreen scores and key environmental indicators"
) %>%
cols_label(
county = "County",
mean_ces_4_0_score = "CES 4.0 (Mean)",
median_pm2_5 = "PM2.5 (Median, µg/m³)",
median_poverty = "Poverty (Median, %)",
age_adjusted_ed_visit_rate = "Asthma ED Rate (Age-Adjusted)"
) %>%
fmt_number(
columns = where(is.numeric),
decimals = 2
) %>%
tab_source_note(
md("*Note:* Asthma ED rate is age-adjusted (CHHS). Environmental indicators are tract-level medians aggregated to county.")
)

```

**Interpretation:** Counties with the highest age-adjusted asthma ED visit rates often also have higher CES scores, higher PM2.5, and higher poverty. This pattern suggests that both environmental and socioeconomic factors may contribute to asthma burden\

## **Stratified table: asthma by CES category**

```{r}
ces_summary <- df_combined %>%
group_by(ces_category) %>%
summarise(
mean_ed_rate  = mean(age_adjusted_ed_visit_rate, na.rm = TRUE),
mean_pm2_5    = mean(median_pm2_5, na.rm = TRUE),
mean_poverty  = mean(median_poverty, na.rm = TRUE),
.groups = "drop"
)

ces_summary %>%
gt() %>%
tab_header(
title = "Asthma ED Rates and Environmental Indicators by CES Category"
) %>%
cols_label(
ces_category = "CES Category",
mean_ed_rate = "Mean Asthma ED Rate (Age-Adjusted)",
mean_pm2_5   = "Mean PM2.5 (Median, µg/m³)",
mean_poverty = "Mean Poverty (Median, %)"
) %>%
fmt_number(columns = where(is.numeric), decimals = 2)
```

**Interpretation:** High-CES counties have higher mean asthma ED rates as well as higher mean PM2.5 and poverty compared to Low-CES counties, supporting the idea that combined environmental and social vulnerabilities are associated with worse asthma outcomes.

### Create `ed_visits_per_100k` as a renamed version of that rate for visualizations

```{r}
df_combined <- df_combined %>%
  mutate(ed_visits_per_100k = age_adjusted_ed_visit_rate)

```

###  Plot 1: Asthma vs CES

```{r Angela visual, echo = TRUE}

ggplot(df_combined, aes(x = mean_ces_4_0_score,
                        y = ed_visits_per_100k)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Relationship Between CalEnviroScreen Score and Asthma ED Visit Rates",
    subtitle = "County-level scores and 2020 asthma emergency visit rates",
    x = "Mean CalEnviroScreen 4.0 Score",
    y = "ED Visit Rate (per 100,000)",
    caption = "Data: OEHHA CalEnviroScreen 4.0 and HCAI Asthma ED Visits (2020)"
  )


```

Or the below includes county names.

```{r}
library(ggplot2)
library(ggrepel)

ggplot(df_combined, aes(x = mean_ces_4_0_score, y = ed_visits_per_100k)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel(aes(label = county), size = 3, max.overlaps = 10) +
  labs(
    title = "County-level Asthma ED Rates vs CES Score",
    subtitle = "Each point represents a California county (2020)",
    x = "Mean CalEnviroScreen 4.0 Score",
    y = "ED Visit Rate (per 100,000)",
    caption = "Data: OEHHA CalEnviroScreen 4.0 and HCAI Asthma ED Visits"
  ) +
  theme_minimal()
```

Interpretation: The positive slope suggests that counties with higher CES 4.0 scores tend to have higher asthma ED visit rates, indicating a possible link between cumulative environmental/social burden and asthma.

#### County-level summary table

-   **Purpose:** Show key county metrics side by side for easy comparison.

-   **Interpretation:** Lets you see which counties have both high environmental exposure and high asthma rates. Could help prioritize intervention targets.

```{r county level table, echo = TRUE}

df_combined %>%
  select(county, mean_ces_4_0_score, ces_category, median_pm2_5, median_diesel_pm, median_poverty, ed_visits_per_100k) %>%
  arrange(desc(ed_visits_per_100k)) %>%
  kable(caption = "County-level Environmental Measures and Asthma ED Rates (2020)")

```

b)  Stratified summary table Purpose: Compare asthma rates by poverty level or CES category. Example: Mean asthma ED rate by CES category:

    Interpretation: Shows if high CES or high poverty counties tend to have higher asthma ED rates.

```{r, echo = TRUE}
df_combined %>%
  group_by(ces_category) %>%
  summarise(mean_ed_rate = mean(ed_visits_per_100k, na.rm = TRUE),
            mean_pm2_5 = mean(median_pm2_5, na.rm = TRUE),
            mean_poverty = mean(median_poverty, na.rm = TRUE))

```

```{r, echo = TRUE}
df_combined %>%
  group_by(county, ces_category) %>%
  summarise(mean_ed_rate = mean(ed_visits_per_100k, na.rm = TRUE),
            mean_pm2_5 = mean(median_pm2_5, na.rm = TRUE),
            mean_poverty = mean(median_poverty, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(mean_ed_rate))
```

### Plot 2: Boxplot of asthma ED rates by CES category

-   **Purpose:** Visually compare distributions across “Low,” “Moderate,” and “High” CES counties.

-   **Interpretation:** The distribution of asthma ED rates is shifted higher in High CES counties compared to Low CES counties, highlighting disparities in asthma burden across environmental risk levels.

```{r, echo = TRUE}
ggplot(df_combined, aes(x = ces_category, y = ed_visits_per_100k, fill = ces_category)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Asthma ED Rates by CES Category",
    x = "CES Category",
    y = "ED Visit Rate (per 100,000)",
    caption = "Data: CalEnviroScreen 4.0 and HCAI Asthma ED Visits (2020)"
  ) +
  theme_minimal()
```

### Plot 3: Asthma ED Rates vs. Median Poverty by County

```{r, echo = TRUE}
  ggplot(df_combined, aes(x = median_poverty, y = ed_visits_per_100k)) +
  geom_point(color = "darkgreen", size = 3) +
    geom_text_repel(aes(label = county), size = 3, max.overlaps = 10)+
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Asthma ED Rates vs. Median Poverty by County",
    x = "Median Poverty Rate (%)",
    y = "ED Visit Rate (per 100,000)",
    caption = "Data: CalEnviroScreen 4.0 and HCAI Asthma ED Visits (2020)"
  ) +
  theme_minimal()
```

**Interpretation**: Counties with higher poverty rates often exhibit higher asthma ED visit rates, suggesting that socioeconomic factors may contribute independently to asthma burden.

### **Plot 4: Scatterplot- Asthma ED Rate vs Median PM2.5**

```{r}
ggplot(df_combined, aes(x = median_pm2_5, y = ed_visits_per_100k)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  geom_text_repel(aes(label = county), size = 3, max.overlaps = 10) +
  labs(
    title = "County-level Asthma ED Rates vs PM2.5",
    x = "Median PM2.5 (µg/m³)",
    y = "ED Visit Rate (per 100,000)",
    caption = "Data: CalEnviroScreen 4.0 and HCAI Asthma ED Visits (2020)"
  ) +
  theme_minimal()
```

**Interpretation:** Shows how air pollution (PM2.5) relates to asthma ED rates.
