---
title: "Milestone 2"
format: pdf
editor: visual
---

```{r}

install.packages("readxl")
library(readxl)
library(janitor)
library(dplyr)


```


Description of dataset:

What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
How does the dataset relate to the group problem statement and question?


5 columns 
- chhs_asthma_ed: county, ED visit rate (and # ED visits?)
- ces_measures: county, asthma, PM2.5, diesel.PM, poverty, pollution burden. 


Questions: 
- in ces_measures, what does the asthma percentile column represent? numbers seem high to be a rate of the general population. 
- also in ces_measures, what the different census tracts mean, and can/should we combine into just county-specific numbers, should we use the mean?  
- do any columns need to be converted to a different type?

cleaning: 
- NA values from ces_measures: asthma, number_of_ed_visits, poverty  Replace with 0, "Unknown", or median(df$column1, na.rm = TRUE) depending on context.

Data elements that need to be converted to a different type
Data elements that need cleaning or re-categorization
Data elements that may be used for future joins (i.e. state, county)

```{r}

ces_datadictionary <- read_excel("~/PHW251_2025/251-Group11/calenviroscreen_datadictionary.xlsx")
  
head(ces_datadictionary)
```

You can add options to executable code like this

```{r}
ces_measures <- read.csv("~/PHW251_2025/251-Group11/calenviroscreen_measures_2021.csv")

ces_measures <- clean_names(ces_measures)

#this is to show the type of each element
str(ces_measures%>% select(california_county, asthma, pollution_burden_score, pm2_5, diesel_pm, poverty))

head(ces_measures)
```

```{r}
#showing the summary statistics
summary(ces_measures$asthma)

```

```{r}

unique(ces_measures$california_county)
```

```{r}
summary(ces_measures$poverty)
```

```{r}
summary(ces_measures$diesel_pm)
```

```{r}
summary(ces_measures$pm2_5)
```

```{r}
summary(ces_measures$pollution_burden_score)
```


```{r}
summary(chhs_asthma_ed$number_of_ed_visits)
```


The `echo: false` option disables the printing of code (only output is displayed).

```{r}
ces_scores_demog <- read.csv("~/PHW251_2025/251-Group11/calenviroscreen_scores_demog_2021.csv")

ces_scores_demog <- clean_names(ces_scores_demog)

head(ces_scores_demog)
```
```{r}
chhs_asthma_ed <- read.csv("~/PHW251_2025/251-Group11/chhs_asthma_ed.csv")

chhs_asthma_ed <- clean_names(chhs_asthma_ed)
str(chhs_asthma_ed%>% select(county, number_of_ed_visits, age_adjusted_ed_visit_rate))


head(chhs_asthma_ed)
```
```{r}

```

